---
title: "A Basic Spatial Planning With Zones Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Basic Spatial Planning With Zones Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  cache = FALSE,
  message = FALSE,
  eval = TRUE
)
```

```{r klippy, eval=TRUE, echo=FALSE, include=TRUE}
# klippy::klippy(position = c('top', 'right'))
```

```{r setup}
library(spatialplanr)
```

# Overview

This code has been written to simplify the process for running a _prioritizr_ analysis on a given region. It is still a work in progress so feel free to submit pull requests with new features and code improvements.

The workflow shown here was based on the [`prioritizr` introduction to zones](https://prioritizr.net/articles/management_zones_tutorial.html).

## Set user parameters 
You can set a region if it is defined in `splnr_get_PlanningUnits`. 
```{r}
Region <- "Coral Sea" # "Australia"
Type <- "Oceans" # "EEZ"
```

Set the shape and area of the planning unit in square kilometres.
```{r}
Shape <- "Hexagon" # "Shape of PUs
PU_size <- 10000 # km2
```

Choose CRS for analysis
```{r}
cCRS <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m no_defs" # Mollweide
```

## Analysis Region

Start your analysis by defining your region and setting up the planning units.

Get the boundary for your chosen region.
```{r}
Bndry <- splnr_get_Boundary(Region, Type, cCRS)

landmass <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>%
  sf::st_transform(cCRS)
```

## Create Planning Units 
```{r}
PUs <- splnr_get_PlanningUnits(Bndry, landmass, PU_size, Shape)
```

We can also use a customised `ggplpot` theme that can be passed as a list to `gg_add()` and that can then be used for all plots. For example:

```{r}
splnr_theme <- list(ggplot2::theme_bw(),
                  ggplot2::theme(
                    legend.position = "right",
                    legend.direction = "vertical",
                    text = ggplot2::element_text(size = 9, colour = "black"),
                    axis.text = ggplot2::element_text(size = 9, colour = "black"),
                    plot.title = ggplot2::element_text(size = 9),
                    axis.title = ggplot2::element_blank()))
```

## Get the features 

For our example, we will use a small subset of charismatic megafauna species of the Coral Sea to inform the conservation plan. We filtered the Aquamaps (Aquamaps.org) species distribution models for our study area for the following species: 
```{r}
Dict <- tibble::tribble(
  ~Common, ~Scientific, ~Category, ~Class,
  "Green sea turtle", "Chelonia_mydas", "Reptiles", "important",
  "Loggerhead sea turtle", "Caretta_caretta", "Reptiles", "important",
  "Hawksbill sea turtle", "Eretmochelys_imbricata", "Reptiles", "important",
  "Olive ridley sea turtle", "Lepidochelys_olivacea", "Reptiles", "important",
  "Saltwater crocodile", "Crocodylus_porosus", "Reptiles", "important",
  "Humpback whale", "Megaptera_novaeangliae", "Mammals", "representative",
  "Common Minke whale", "Balaenoptera_acutorostrata", "Mammals", "representative",
  "Dugong", "Dugong_dugon", "Mammals", "important",
  "Grey nurse shark", "Carcharias_taurus", "Sharks and rays", "important",
  "Tiger shark", "Galeocerdo_cuvier", "Sharks and rays", "important",
  "Great hammerhead shark", "Sphyrna_mokarran", "Sharks and rays", "representative",
  "Giant oceanic manta ray", "Mobula_birostris", "Sharks and rays", "important",
  "Reef manta ray", "Mobula_alfredi", "Sharks and rays", "important",
  "Whitetip reef shark", "Triaenodon_obesus", "Sharks and rays", "important",
  "Red-footed booby", "Sula_sula", "Birds", "important"
)
```

These species were not chosen based on their importance for this region and only represent an example for visualization purposes. 

**Note:** The structure of the `tribble`above is required for some of the downstream plotting. *Common* denotes the common name of a species, *Scientific* the scientific name in the format used by Aquamaps, *Category* is the category that a species belongs to and *Class* represents the importance of the species for the conservation plan.

### Convert the probabilities to binary data and feature names

```{r}
datEx_species_bin <- spDataFiltered %>%
   splnr_apply_cutoffs(Cutoffs = 0.5)

col_name <- spDataFiltered %>%
  sf::st_drop_geometry() %>%
  dplyr::select(-"cellID") %>%
  colnames()
```

### Spatial planning including zones

So far, all steps were exactly the same as in a spatial plan that does not include zones. To have a direct comparison, we will first run a spatial plan that does not include zoning. For this, we will define targets for our features (here: 30% for all features) and create a conservation problem where we include an area-based cost in our prioritisation.

```{r}
target <- rep(0.3, nrow(Dict))

p1 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397)),
                                     col_name,
                                     "Cost1") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_relative_targets(target) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(verbose = FALSE)
```

We then solve the problem and plot the solution.

```{r, fig.width=9}
s1 <- p1 %>%
  prioritizr::solve.ConservationProblem()

(ggSoln <- splnr_plot_Solution(s1) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))
```

We can also have a look at how well the set target was met.

```{r}
s1T <- s1 %>%
  dplyr::select(tidyselect::starts_with(c("solution"))) %>%
  sf::st_drop_geometry() %>%
  tibble::as_tibble()

r1 <- prioritizr::eval_feature_representation_summary(p1, s1T)
print(r1)
```

We will now look how this spatial plan needs to be extended to include multiple management zones by using `zones()`.

### Different cost layers

We first assume that we have to zones of interest that have the same set of features, but separate cost layers. For this, we need to set targets for both zones, in our case 20% for all features in the first zone and 5% of all features in the second zone. 

```{r}
target2 <- matrix(NA, ncol = 2, nrow = nrow(Dict))
target2[, 1] <- 0.2
target2[, 2] <- 0.05
```

We then create a `zones` object where we specify the number of zones and the features in the zone. Here, we define two zones that both have the same features.

```{r}
z2 <- prioritizr::zones("zone 1" = col_name, "zone 2" = col_name)
```

Both zones need a specified cost and targets. We use one area-based cost and one cost layer that uses random numbers, the pre-defined targets and pass the `zones`object to the problem formulation before solving the conservation problem.

```{r}
p2 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                            Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(verbose = FALSE)

s2 <- p2 %>%
  prioritizr::solve.ConservationProblem()

```

The solution can be plotted using the function `splnr_plot_SolutionZones()` that works identical the function `splnr_plot_Solution()` but for the output of a solution generated with zones.

```{r, fig.width = 9}
(gg_s2 <- splnr_plot_SolutionZones(s2) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))
```

We can see that we now how planning units that are not selected, and planning units that have either been selected for zone 1 or zone 2.

### Boundary penalty

Similar to "normal" priritisations, zones also allows including a boundary penalty. There are several options how this can be done within zones or between zones. 

#### Within the same zone

```{r}
zm1 <- diag(2) 
print(zm1)
```

The zeros in this matrix denote that there is no penalty between zones, whereas ones introduce a boundary penalty when we define the conservation problem. Here, we only introduce penalties within zone 1 AND zone 2 respectively.

```{r}
p3 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                              Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_boundary_penalties(0.5, zone = zm1) %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(time_limit = 10, verbose = FALSE)

s3 <- p3 %>%
  prioritizr::solve.ConservationProblem()

```

With `add_boundary_penalties()` we set a penalty of 0.5 between the zones we defined in `zm1`. For the purpose of this example, we also set a 10 second time limit for the prioritisation. The results show less fragmentation than in the previous solution without penalties:

```{r, fig.width=9}
(gg_s3 <- splnr_plot_SolutionZones(s3) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))

```

A boundary penalty can also be introduced to only one zone:

```{r, fig.width=9}
zm2 <- zm1
zm2[2,2] = 0

p4 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                              Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_boundary_penalties(0.5, zone = zm2) %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(time_limit = 10, verbose = FALSE)

s4 <- p4 %>%
  prioritizr::solve.ConservationProblem()

(gg_s4 <- splnr_plot_SolutionZones(s4) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))
```
 
#### Across zones

Similarly, penalties can be applied between zones if specified in the problem.

```{r}
zm3 <-  matrix(1, ncol = 2, nrow = 2)
print(zm3)
```

```{r, fig.width=9}
p5 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                              Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_boundary_penalties(0.5, zone = zm3) %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(time_limit = 10, verbose = FALSE)

s5 <- p5 %>%
  prioritizr::solve.ConservationProblem()

(gg_s5 <- splnr_plot_SolutionZones(s5) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))

```

### Adding constraints

Another addition often used in conservation planning are constraints. We often lock in or lock out areas of interest for various reasons. In this example, we will lock in already existing MPAs in one of the zones.

When using zones with constraints, we use the `prioritizr` function `add_manual_locked_constraints()` that requires the input to be in a specific format, so we first prepare the MPA data.

```{r}
mpas <- MPAsCoralSea %>%
  dplyr::mutate(zone = "zone 1") %>%
  dplyr::rename(pu = cellID,
                status = wdpa) %>%
  sf::st_drop_geometry() %>%
  tibble::tibble() %>%
  dplyr::filter(status == 1)
```

We need to specify which zone the constraint should be applied to, give information on the planning unit identifier, rename the column of interest to "status" and filter out only the planning units that should be locked in.

```{r}
p6 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                              Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_manual_locked_constraints(mpas) %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(time_limit = 10, verbose = FALSE)

s6 <- p6 %>%
  prioritizr::solve.ConservationProblem()
```

The results can the be visualized again:

```{r, fig.width = 9}
(gg_s6 <- splnr_plot_SolutionZones(s6) +
    gg_add(PUs = PUs, Bndry = Bndry,
           lockedInAreas = MPAsCoralSea, colInterest = MPAsCoralSea$wdpa, Type = "Full", alphaLI = 0.1, colorLI = "yellow",
           land = landmass, cropLand = PUs, ggtheme = splnr_theme))
```

### Objective functions

So far we have only used the minimum set objective function to meet all conservation targets whilst minimizing the cost. However, this is not the only objective function available in `prioritizr`. For example when maximising mangrove ecosystem services, we employed a minimum shortfall objective function instead [Dabalà et al 2022](https://www.researchgate.net/publication/363818810_Priority_areas_to_protect_mangroves_and_maximise_ecosystem_services). For this objective function, we allocate a budget that cannot be exceeded (for example area >30%) whilst still trying to minimise the cost and meeting the targets - but not all targets might be met (shortfall).

Zones also works with objective functions other than the minimum set objective function.

```{r}
p7 <- prioritizr::problem(datEx_species_bin %>% dplyr::mutate(Cost1 = rep(1, 397), #when giving sf input, we need as many cost columns as we have zones
                                                              Cost2 = runif(n = dim(.)[[1]])),
                          z2,
                          cost_column = c("Cost1", "Cost2"))%>%
  prioritizr::add_min_shortfall_objective(c(8, 0.3)) %>%
  prioritizr::add_relative_targets(target2) %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(time_limit = 10, verbose = FALSE)
```

Here we apply `add_min_shortfall_objective()` and provide to separate budgets for the two zones (8 and 0.3). We can visualize the output again after solving the conservation problem.

```{r, fig.width = 9}
s7 <- p7 %>%
  prioritizr::solve.ConservationProblem()

(gg_s7 <- splnr_plot_SolutionZones(s7) +
    gg_add(PUs = PUs, Bndry = Bndry, land = landmass, cropLand = PUs, ggtheme = splnr_theme))

```

