---
title: "Pacific-Centred Spatial Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pacific-Centred Spatial Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE, 
  cache=FALSE, 
  message=FALSE,
  eval=FALSE
)
```

```{r setup}
# library(spatialplanr)
```

For a basic introduction to using `spatialplanr` see the main vignette. This vignette will outline how to undertake a spatial analysis using a Pacific Ocean centred analysis (ie. across the dateline). The same functions are generally used, but additional options may need to specified.

## Set user parameters 
You can set a region if it is defined in `SpatPlan_Get_PlanningUnits`. THe currently defined regions are: `Global`, `Australia`, `WestPacific`. If you need a new region, please submit a PR so others can benefit.
```{r}
Region <- "WestPacific"
# Region <- "Pacific"
```

Set the shape and area of the planning unit in square kilometres.
```{r}
Shape <- "Hexagon" # "Shape of PUs
PU_size <- 2620 # km2 (0.5 deg at equator)
```

Set the depth range and probability cutoff (`CO`) for extracting AquaMaps data.
```{r}
MinDepth <- 0
MaxDepth <- 200
CO <- 0.5 #Cutoff level for AquaMaps
```

Choose CRS for analysis
```{r}
cCRS <- "+proj=robin +lon_0=180 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs" # Robinson: Pacific-centred
```

## Analysis Region

Start your analysis by defining your region and setting up the planning units.

Get the boundary for your chosen region.
```{r}
Bndry <- SpatPlan_Get_Boundary(Region, cCRS)
```

Get the land boundaries to remove overlap. The object "world" represents the land mass that should be removed (inverse = FALSE)
```{r}
world <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(cCRS)
```
If Region = WestPacific, demarcating boundary and removing land areas would be different
The object "world" here represents the planning region (inverse = TRUE)

## Create Planning Units 
```{r}
PUs <- SpatPlan_Get_PlanningUnits(Bndry, world, PU_size, Shape)
```
We can check what the PUs look like
```{r, fig.width=7}
(ggPU <- SpatPlan_Plot_PUs(PUs, world)) # Plot Planning Units
```

## Get the features 

### Aquamaps 
Get aquamaps data for our species and region 
TODO Check the overlay of raw v processed species distribution
If using Pacific-centered projection, define western and eastern limits
west_limit = -120 # use negative numbers for western limit
east_limit = 130 # use positive numbers for eastern limit

```{r}
aqua_sf <- SpatPlan_Get_AquaMaps(PUs, cCRS, MinDepth, MaxDepth, CutOff = CO)
```

### IUCN 
TODO
IUCN <- fSpatPlan_Get_IUCN(PUs, cCRS)

### MICO 
TODO
MICO <- fSpatPlan_Get_MICO(PUs, cCRS)

### Seafloor 
TODO
SeaFloor <- fSpatPlan_Get_SeaFloor(PUs, cCRS)

### Geomorphic features 
TODO
GeoMorph <- fSpatPlan_Get_GeoMorph(PUs, cCRS)

### Longhurst Provinces
You can also assign a Longhurst Province to each PU and set targets individually by province.Here we simply demonstrate this, but do not implement it into a spatial plan.
```{r, fig.width=7}
Longhurst <- SpatPlan_Match_Longhurst(PUs, cCRS)
(ggLong <- SpatPlan_Plot_Longhurst(Longhurst, world))
```

### Get locked in areas 
```{r, fig.width=7}
LockedIn <- SpatPlan_Get_MPAs(PUs, cCRS)
(ggMPA <- SpatPlan_Plot_MPAs(LockedIn, world)) # Plot Locked in areas
```

## Get Fisheries cost 
```{r, fig.width=7}
Cost <- SpatPlan_Get_FishCost(PUs, cCRS)
(ggCost <- SpatPlan_Plot_Cost(Cost, world)) # Plot cost
```

## Set up targets 

### Identical fixed targets 
```{r}
Targets <- 0.2
```

### Inverse area targets 
```{r}
Targets <- SpatPlan_Get_TargetsIA(aqua_sf, minTarget, maxTarget)
```

### Add different targets for IUCN RedList categories 
```{r, eval=FALSE}
speciesCol <- "Species"
Targets <- Targets %>% 
  SpatPlan_Match_IUCNRedList(speciesCol) %>% # Add RL data to the df
  dplyr::mutate(Target = dplyr::case_when(IUCN_Category %in% c("EX","EW","CR","EN","VU") ~ 1,
                                          TRUE ~ Targets$Target))
```

## Set up the planning problem 

### Get the list of features
```{r}
features <- aqua_sf %>% 
  dplyr::as_tibble() %>% 
  dplyr::select(-geometry) %>% 
  names()

out_sf <- cbind(Cost, LockedIn, aqua_sf)

p1 <- prioritizr::problem(out_sf, features, "Cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_relative_targets(Targets$Target) %>%
  prioritizr::add_locked_in_constraints("locked_in") %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(verbose = FALSE)
```

## Solve the planning problem 
```{r}
s1 <- prioritizr::solve(p1) %>% 
  dplyr::mutate(solution_1 = as.logical(solution_1)) # Making it logical helps with the plotting #TODO move this to the plotting section?
```

## Plotting 
```{r, fig.width=7}
(ggSol <- SpatPlan_Plot_Solution(s1, PUs, world)) # Plot Solution

ggFeatureNo <- SpatPlan_Plot_FeatureNo(aqua_sf, world) # Plot Number of features

#+ echo=FALSE, fig.width=12
(gg <- patchwork::wrap_plots(list(ggPU, ggMPA, ggCost, ggFeatureNo, ggSol), ncol = 3))
# ggsave(file.path("Figures", "Final_Solution.pdf"), plot = gg, height = 20, width = 40, units = "cm")
```

