---
title: "A Basic Spatial Planning Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spatialplanr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning=FALSE, 
  cache=FALSE, 
  message=FALSE
)
```

```{r klippy, eval=TRUE, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

```{r setup}
library(spatialplanr)
```

# Overview

This repository is specific to the work of the [https://mathmarecol.github.io](Mathematical Marine Ecology Lab) at the University of Queensland. It is not intended to be used by anyone outside the MME Lab without the permission of Professor Anthony Richardson. 

This code has been written to simplify the process for running a _prioritizr_ analysis on a given region. It is still a work in progress so feel free to submit pull requests with new features and code improvements.

The code depends on `sf`, `terra`, `tidyverse`, `rnaturalearth`, `prioritizr`, `stars`, `patchwork`.    

To use this code, you will need to download and expand `MME1DATA-Q1215/SpatialPlanning/Data.zip` to the directory `GitHub/SpatialPlanning/Data/`. Note that the download is only 2GB, but the expanded data is 35 GB in size. If you need help subsetting the data to your region due to memory or HD space constraints, contact Jason.

If you run into memory problems, you can try increasing the amount of the HD space gdal is allowed to use. 
`Sys.setenv(GDAL_MAX_BAND_COUNT = 120000)`

## Set user parameters 
You can set a region if it is defined in `SpatPlan_Get_PlanningUnits`. 
```{r}
Region <- "Australia" # "South Atlantic Ocean"
Type <- "EEZ" # "Oceans"
```

You can also define a region with square boundaries if you need.
```{r, eval=FALSE}
Region <- c(xmin = 150, xmax = 160, ymin = -40, ymax = -30)
```

Set the shape and area of the planning unit in square kilometres.
```{r}
Shape <- "Hexagon" # "Shape of PUs
PU_size <- 20000 # km2
```

Choose CRS for analysis
```{r}
cCRS <- "+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m no_defs" # Mollweide
```

## Analysis Region

Start your analysis by defining your region and setting up the planning units.

Get the boundary for your chosen region.
```{r}
Bndry <- SpatPlan_Get_Boundary(Region, Type, cCRS)
```

Get the land boundaries to remove overlap. The object "landmass" represents the land mass that should be removed (inverse = FALSE)
```{r}
landmass <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(cCRS)
```

## Create Planning Units 
```{r}
PUs <- SpatPlan_Get_PlanningUnits(Bndry, landmass, PU_size, Shape)
```

We can check what the PUs look like
```{r, fig.width=7}
(ggPU <- SpatPlan_Plot_PUs(PUs, landmass)) # Plot Planning Units
```

## Get the features 

### Aquamaps 
Get aquamaps data for our species and region 

Set the depth range and probability cutoff (`CO`) for extracting AquaMaps data.
```{r}
MinDepth <- 0
MaxDepth <- 200

AQM <- SpatPlan_Get_AquaMaps(PUs, MinDepth, MaxDepth)
```

### Geomorphic features 
Add geomorphic features to the existing features object.
```{r}
GeoMorph <- SpatPlan_Get_Geomorphic(PUs, Feature = c("Shelf", "Ridge", "Plateau", "Seamount"))
```


### Apply cutoffs
Add geomorphic features to the existing features object.
```{r}
Cutoffs <- 0.5
AQM_bin <- SpatPlan_Apply_Cutoffs(AQM, Cutoffs)

Cutoffs <- c(Shelf = 0.01, Ridge = 0.2, Plateau = 0.1, Seamount = 0.3)
GeoMorph_bin <- SpatPlan_Apply_Cutoffs(GeoMorph, Cutoffs)
```



### Join all features together.
```{r}
features <- GeoMorph %>% 
  dplyr::left_join(AQM %>% sf::st_drop_geometry(), by = "cellID")

(ggFeatureNo <- SpatPlan_Plot_FeatureNo(features, landmass)) # Plot Number of features
```

### Longhurst Provinces
You can also assign a Longhurst Province to each PU and set targets individually by province. Here we simply demonstrate this, but do not implement it into a spatial plan.
```{r, fig.width=7}
Longhurst <- SpatPlan_Match_Longhurst(PUs)
(ggLong <- SpatPlan_Plot_Longhurst(Longhurst, landmass))
```

### Get locked in areas 
```{r, fig.width=7}
LockedIn <- SpatPlan_Get_MPAs(PUs, "Australia") %>% 
  SpatPlan_Apply_Cutoffs(0.5) %>% 
  dplyr::mutate(wdpa = as.logical(wdpa))
(ggMPA <- SpatPlan_Plot_MPAs(LockedIn, landmass)) # Plot Locked in areas
```

## Get Fisheries cost 
```{r, fig.width=7}
Cost <- SpatPlan_Get_FishCost(PUs, group = "all")
(ggCost <- SpatPlan_Plot_Cost(Cost, landmass)) # Plot cost
```

## Set up targets 

### Identical fixed targets 
```{r}
Targets <- data.frame(Target = 0.2)
```

### Inverse area targets 
Set minimum and maximum limits for inverse area targets
```{r, eval = FALSE}
minTarget = 0.2
maxTarget = 0.5

Targets <- SpatPlan_Get_TargetsIA(features, minTarget, maxTarget)
```

### Add different targets for IUCN RedList categories 
To run this code you will need to sign up for your own IUCN RedList token. Start by running `rredlist::rl_use_iucn()`. 
```{r, eval=FALSE}
speciesCol <- "Species"
Targets <- Targets %>% 
  SpatPlan_Match_IUCNRedList(speciesCol) %>% # Add RL data to the df
  dplyr::mutate(Target = dplyr::case_when(IUCN_Category %in% c("EX","EW","CR","EN","VU") ~ 1,
                                          TRUE ~ Targets$Target))
```

## Set up the planning problem 

### Get the list of features
```{r}
feature_names <- features %>% 
  sf::st_drop_geometry() %>% 
  names()

out_sf <- cbind(Cost, 
                LockedIn %>% sf::st_drop_geometry(), 
                features %>% sf::st_drop_geometry())

p1 <- prioritizr::problem(out_sf, feature_names, "Cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_relative_targets(Targets$Target) %>%
  prioritizr::add_locked_in_constraints("wdpa") %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(verbose = FALSE)
```

## Solve the planning problem 
```{r}
s1 <- prioritizr::solve(p1)
```

## Plotting 
```{r, fig.width=7}
(ggSol <- SpatPlan_Plot_Solution(s1, PUs, landmass)) # Plot Solution

gg <- patchwork::wrap_plots(list(ggPU, ggMPA, ggCost, ggFeatureNo, ggSol), ncol = 3)
```

## Compare 2 solutions
Here we run another problem/solution combo and compare the output with our original one. With a much lower target (30%), we expect less area set aside.

### Create the second planning problem 
```{r}
p2 <- prioritizr::problem(out_sf, feature_names, "Cost") %>%
  prioritizr::add_min_set_objective() %>%
  prioritizr::add_relative_targets(0.5) %>%
  prioritizr::add_locked_in_constraints("wdpa") %>%
  prioritizr::add_binary_decisions() %>%
  prioritizr::add_default_solver(verbose = FALSE)
```

### Solve the planning problem 
```{r}
s2 <- prioritizr::solve(p2)
```

### Plot comparison
```{r, fig.width=7}
gg <- SpatPlan_Plot_Comparison(s1, s2, landmass)
```
